---
title: "Data cleaning and auditing"
format: 
  html:
    toc: true
    toc-location: right
    smooth-scroll: true
    html-math-method: katex
    code-fold: false
self-contained: true
editor: source
author: 'Florencia Grattarola'
date: "`r format(Sys.time(), '%Y-%m-%d')`"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: libraries
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(httr)
library(jsonlite)
library(countrycode)
library(janitor)
library(readxl)
library(sf)
sf_use_s2(FALSE)
library(tmap)
tmap_mode('view')
library(tidyverse)
options(knitr.kable.NA = '')
```

## Data cleaning

### Read data

```{r}
#| label: read-data
#| echo: true
#| eval: true
#| message: false
#| warning: false

raw_metadata <- read_xlsx('data/draft/metada_work_version.xlsx', guess_max = 4000) 
```

### Check columns

```{r}
#| label: check-col-names
#| echo: true
#| eval: true
#| message: false
#| warning: false

raw_metadata <- raw_metadata %>% 
  janitor::clean_names() %>% 
  janitor::remove_empty(c('rows', 'cols')) 

```

### Check `source` fields

The fields are: `name_orig`, `format`, and `language`.


  - Make sure there are no `\r`, `\n`, other weird characters or typos. 
  - Capitalise and clean `language`.  

```{r}
#| label: check-source-fields
#| echo: true
#| eval: false
#| message: false
#| warning: false

# name
raw_metadata %>% 
  filter(grepl('http', name_orig)) %>% 
  group_by(name_orig) %>% count

raw_metadata %>% 
  filter(is.na(name_orig)) %>% 
  distinct(url_clean)

raw_metadata %>% 
  mutate(name_orig = str_squish(name_orig)) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\\\\")) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\"")) %>% 
  group_by(name_orig) %>% 
  summarise(n_countries = n_distinct(country)) %>% 
  select(name_orig, n_countries) %>% 
  print(n=10)

# format
raw_metadata %>% 
  mutate(format = ifelse(format == 'NA', NA, str_squish(format))) %>% 
  distinct(format)

# language
raw_metadata %>% 
  mutate(language = ifelse(language == 'NA', NA, str_squish(language))) %>% 
  mutate(language = str_replace(language, "/|\\|", ' | ')) %>% 
  distinct(language) %>% 
  print(n=50)
```

#### DOUBTS

  - [ ] There's one source that has a URL instead of a name, that includes 12 records. 
  
  The `name_orig` is: `https://www.odonat-grandest.fr/listes-rouges-grand-est-etat-avancement/`.  
  The `id` values are: `[1] 446 447 448 449 450 451 452 453 454 455 456 457`.   

  - [ ] There are some `formats` with `?` as value. What should they be?

#### FIX

```{r}
#| label: check-col-names-fix
#| echo: true
#| eval: false
#| message: false
#| warning: false

# French: Liste rouge des Amphibiens du Grand Est
# English: Red list of Mammals of Grand Est    

raw_metadata %>% 
  mutate(name_orig = ifelse(grepl('htt', name_orig) & 
                               state_province == 'Grand Est',
                             str_glue('Red list of {group} of Grand Est'), name_orig)) %>% 
  mutate(name_orig = case_when(grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Birds' ~ 
                                 'Liste rouge des Oiseaux du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Mammals' ~ 
                                 'Liste rouge des XXX du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Orthoptera' ~ 
                                 '',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Fishes' ~ 
                                 '',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Butterflies' ~ 
                                 '',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Night butterflies' ~ 
                                 '',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Ladybugs' ~ 
                                 '',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Branchiopoda' ~ 
                                 '',
                               .default = name_orig))
  filter(state_province == 'Grand Est') %>% select(group, name_orig)

raw_metadata %>% 
  mutate(language = ifelse(language == 'NA', NA, str_squish(language))) %>% 
  mutate(language = str_replace(language, "/|\\|", ' | ')) %>% 
  distinct(language)

raw_metadata %>% 
  mutate(format = ifelse(format == '?', NA, format)) %>% 
  distinct(format)

raw_metadata %>% 
  mutate(name_orig = str_squish(name_orig)) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\\\\")) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\"")) 
```

### Check `Location` fields

The fields are: `continent`, `country`, `state_province`, `gadm_level_1`, `gadm_level_2`, `region_custom`, `region_detail`, and `iso_2`.

  - Make sure there are no typos, thus duplicates.   
  - Capitalise `continent`, `country`, `state_province` names.  
  - Check ISO codes.  
  - Check GADM levels.  
  - Clean `region_custom` and `region_detail`.  
  
```{r}
#| label: check-location-fields
#| echo: true
#| eval: false
#| message: false
#| warning: false

# check continent
raw_metadata %>% 
  mutate(continent = str_squish(str_replace_all(str_squish(continent), '\\|', ' | '))) %>%
  mutate(continent = str_replace_all(continent, '_', ' ')) %>%
  mutate(continent = str_to_title(continent)) %>% 
  distinct(continent)

# check country
raw_metadata %>% 
  mutate(country = ifelse(country == 'NA', NA, str_squish(country))) %>% 
  mutate(country = str_replace_all(country, '_', ' ')) %>%
  mutate(country = ifelse(country == 'USSR', NA, str_to_title(country))) %>%
  mutate(country = str_replace_all(country, 'And ', 'and ')) %>% 
  mutate(country = str_replace_all(country, 'Of', 'of')) %>% 
  mutate(country = str_replace_all(country, 'The', 'the')) %>% 
  distinct(country)

# check state_province
raw_metadata %>% 
  mutate(state_province = ifelse(state_province == 'NA', NA, str_squish(state_province))) %>%
  mutate(state_province = str_to_title(state_province)) %>% 
  mutate(state_province = str_replace_all(state_province, 'And ', 'and ')) %>% 
  mutate(state_province = str_replace_all(state_province, 'Of', 'of')) %>% 
  mutate(state_province = str_replace_all(state_province, 'The', 'the')) %>% 
  filter(!is.na(state_province)) %>% 
  filter(state_province != gadm_level_1) %>% 
  distinct(country, state_province, gadm_level_1, iso_2, iso_3) %>% 
  print(n=100)

# check gadm_level_1 and gadm_level_2
raw_metadata %>% 
  mutate(gadm_level_1 = ifelse(gadm_level_1 == 'NA', NA, str_squish(gadm_level_1))) %>%
  mutate(gadm_level_1 = str_to_title(gadm_level_1)) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'And ', 'and ')) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'Of', 'of')) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'The', 'the')) %>% 
  filter(!is.na(gadm_level_1)) %>% distinct(gadm_level_1) %>% 
  arrange(gadm_level_1) %>% print(n=100)

raw_metadata %>% 
  mutate(gadm_level_2 = ifelse(gadm_level_2 == 'NA', NA, str_squish(gadm_level_2))) %>%
  mutate(gadm_level_2 = str_to_title(gadm_level_2)) %>% 
  filter(!is.na(gadm_level_2)) %>% select(gadm_level_2)

# check region_custom and region_detal
raw_metadata %>% 
  mutate(region_custom = ifelse(region_custom == 'NA', NA, str_squish(region_custom))) %>% 
  mutate(region_detail = ifelse(region_detail == 'NA', NA, str_squish(region_detail))) %>% 
  mutate(region_detail = str_squish(str_replace_all(str_squish(region_detail), '\\|', ' | '))) %>%
  filter(!is.na(region_custom)) %>%
  distinct(region_custom, region_detail, iso_2, iso_3) %>% 
  print(n=100)

# check iso_2 and iso_3
raw_metadata %>% 
  janitor::clean_names() %>% 
  janitor::remove_empty(c('rows', 'cols')) %>% 
  mutate(iso_2 = ifelse(iso_2 == 'NA' & country != 'Namibia', NA, str_squish(iso_2))) %>% 
  mutate(iso_2 = str_squish(str_replace_all(str_squish(iso_2), '\\|', ' | '))) %>%
  mutate(iso_3 = ifelse(iso_3 == 'NA', NA, str_squish(iso_3))) %>% 
  mutate(iso_3 = str_squish(str_replace_all(str_squish(iso_3), '\\|', ' | '))) %>%
  select(country, iso_2, iso_3, region_custom, region_detail) %>%  
  filter(is.na(iso_2)) #%>% distinct()



```

#### DOUBTS

  - [ ] What to do with `country == USSR`? Should be custom region and not country.  
  - [x] Kosovo is `iso_2 = XK`? Yes, it's a temporary, unofficial country code top-level domain for Kosovo.    
  - [x] Is `Antartica` and `French Souther Territories` correct? Yes.    
  - [ ] Lots of `state_province` and `GADM_level_1` to correct. Is GADM_1 in English? e.g., 'Südtirols' = 'Bolzano' (state_name in German?), 'Centre-Val De Loire' = 'Centre' (something missing?), 'Corsica' = 'Corse'  (something missing?), 'Flanders' = 'Vlaanderen' (English?), 'Opole Voivodeship' = 'Opole'  (something missing?).  
    

### Check `Taxon` fields

The fields are: `kingdom`, `phylum`, `subphylum`, `class`, `order`, and `group`.

  - Check duplicates in all fields.
  - Rename `group` to `taxa`.

I searched names using the GBIF backbone, using my own custom function `nameMatcherGBIF()`.  
 
```{r}
#| label: function-nameMatcherGBIF
#| echo: true
#| eval: true
#| message: false
#| warning: false

# gbif name parser
nameMatcherGBIF <- function(sp_name_list) {
  
  # api <- 'http://api.gbif.org/v1/parser/name'
  api <- 'http://api.gbif.org/v1/species/match'
  
  name_parsed <- tibble(sp_name = character(),
                        scientificName = character(),
                        kingdom = character(),
                        phylum = character(),
                        class = character(),
                        order = character(),
                        family = character(),
                        genus = character(),
                        specificEpithet = character(),
                        species = character(),
                        status = character(), 
                        rank = character())
  
  for(sp_name in sp_name_list){
    # cat(sp_name, '\n')
    call_url <- str_glue('{api}?name={sp_name}&strict=true&verbose=false')
    get_json_call <- GET(url = URLencode(call_url)) %>%
      content(as = "text") %>% fromJSON(flatten = TRUE)
    
    if(get_json_call$matchType == 'NONE') {
      name_parsed_i <- tibble(sp_name = sp_name,
                              scientificName = NA,
                              kingdom = NA,
                              phylum = NA,
                              class = NA,
                              order = NA,
                              family = NA,
                              genus = NA,
                              specificEpithet = NA,
                              species = NA,
                              status = NA, 
                              rank = NA)
      name_parsed <- rbind(name_parsed, name_parsed_i)
    } else{
      name_parsed_i <- tibble(sp_name = sp_name,
                              scientificName = ifelse(exists('scientificName',get_json_call), get_json_call$scientificName, NA),
                              kingdom = ifelse(exists('kingdom',get_json_call), get_json_call$kingdom, NA),
                              phylum = ifelse(exists('phylum',get_json_call), get_json_call$phylum, NA),
                              class = ifelse(exists('class',get_json_call), get_json_call$class, NA),
                              order = ifelse(exists('order',get_json_call), get_json_call$order, NA),
                              family = ifelse(exists('family',get_json_call), get_json_call$family, NA),
                              genus = ifelse(exists('genus',get_json_call), get_json_call$genus, NA),
                              specificEpithet = ifelse(exists('specificEpithet',get_json_call), get_json_call$specificEpithet, NA),
                              species = ifelse(exists('species',get_json_call), get_json_call$species, NA),
                              status = ifelse(exists('status',get_json_call), get_json_call$status, NA), 
                              rank = ifelse(exists('rank',get_json_call), get_json_call$rank, NA))
      
      name_parsed <- rbind(name_parsed, name_parsed_i)
    }
  }
  return(name_parsed)
}

sp_list <- raw_metadata %>% 
  distinct(kingdom, phylum, subphylum, class, order, group) %>% 
  mutate(group = str_trim(group)) %>% 
  mutate(group = str_to_title(group)) %>% 
  mutate(group = str_replace_all(group, 'And ', 'and ')) %>% 
  mutate(group = str_replace_all(group, 'Of', 'of')) %>% 
  mutate(group = str_replace_all(group, 'The', 'the')) %>%
  pull(group)

sp_list_matched <- nameMatcherGBIF(sp_list) %>% suppressMessages()

sp_list_matched <- sp_list_matched %>% 
  mutate(scientificName = case_when(grepl('flora', sp_name, ignore.case=T) ~ 'Plantae',
                                    grepl('fauna', sp_name, ignore.case=T) ~ 'Animalia',
                                    grepl('tunicata', sp_name, ignore.case=T) ~ NA,
                                    .default = scientificName)) %>% 
  mutate(kingdom = case_when(grepl('flora', sp_name, ignore.case=T) ~ 'Plantae',
                             grepl('fauna', sp_name, ignore.case=T) ~ 'Animalia',
                             grepl('tunicata', sp_name, ignore.case=T) ~ NA,
                             .default = kingdom)) %>% 
  mutate(phylum = ifelse(phylum == 'chordata', 'Chordata', phylum))

sp_list_unmatched <- sp_list_matched %>% 
  filter(is.na(scientificName)) %>% pull(sp_name)

sp_list_matched %>% filter(!is.na(kingdom)) %>% nrow() # matched
length(sp_list_unmatched) # not matched
```

When the taxon name (i.e., `group`) was not found I kept the previous taxonomic fields' values.

```{r}
#| label: check-taxon-fields
#| echo: true
#| eval: true
#| message: false
#| warning: false

merged_list <- left_join(raw_metadata %>%
                           mutate(group = str_trim(group)) %>%
                           distinct(group) %>%
                           arrange(group),
                         sp_list_matched %>%
                           filter(!is.na(scientificName)) %>%
                           rename(group=sp_name) %>% distinct()) %>%
  arrange(group)

raw_metadata_taxon_list <- raw_metadata %>%
  mutate(group = str_trim(group)) %>% 
  distinct(group, .keep_all = T) %>% 
  select(kingdom, phylum, subphylum, class, order, group) %>% 
  arrange(group)

left_join(raw_metadata %>% mutate(group = str_trim(group)) %>%
            select(-c(kingdom,phylum,subphylum,class,order)),
          bind_rows(merged_list %>% 
                      filter(!is.na(scientificName)),
                    merged_list %>% 
                      filter(is.na(scientificName)) %>%
                      select(group) %>% 
                      left_join(. , raw_metadata_taxon_list))) %>% 
  mutate(group = str_trim(group)) %>% 
  mutate(group = str_to_title(group)) %>% 
  mutate(group = str_replace_all(group, 'And ', 'and ')) %>% 
  mutate(group = str_replace_all(group, 'Of', 'of')) %>% 
  mutate(group = str_replace_all(group, 'The', 'the')) %>%
  mutate(group = str_replace_all(group, 'Et Al.', 'et al.')) %>%
  distinct(group, kingdom, phylum, class, order, family, rank) %>% 
  arrange(kingdom, phylum, class, order) %>% 
  kableExtra::kbl(booktabs = T) %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))
```

  - [ ] The value `Tracheophyta` is for vascular plants. Some groups with value `Flora` have this but some others don't. Is it correct?
  - [ ] There are some groups with value `Fauna` that have  `Chordata` and `Vertebrata` as `phylum` and `subphylum`, but some others don't. Is it correct?
  - [ ] The group `Onychophora` should be phylum `Onychophora`.  
  - [ ] There's a typo on group `Opilioness`, should be `Opiliones`.  
  - [ ] Correct the value to order `Pseudoscorpiones` for the group `False Scorpions`.  
  - [ ] The group `Entomostraca` should not have `Copepoda` as class.
  - [ ] Crabs should have class `Malacostraca`.  
  - [ ] Sharks should not have `Chondrichthyes` as class.  
  - [ ] Groups `Cetartiodactyla` and `Marine Cetartiodactyla` should be order `Artiodactyla`.  
  - [ ] Carnivores should be order `Carnivora`.
  - [ ] Group `Lizards` should be order `Squamata`.  
  - [ ] Group `Lampreys` should be class `Petromyzonti` and order `Petromyzontiformes`.
  - [ ] There's a typo on group `Musslels`, should be `Mussels`.  
  - [ ] There's a typo on the phylum `chordata`, should be uppercase.
  - [ ] The groups `Orchids` and `Bromeliads`, are missing phylum, class and family. They should be `Tracheophyta` and `Liliopsida`. Also family `Orchidaceae` and  `Bromeliaceae`.   
  - [ ] The group `Wild Cinnamon`, is missing phylum, class and order. They shoud be `Tracheophyta` `Magnoliopsida` `Canellales` `Canellaceae`, and the class should have family `Canellaceae`.
  - [ ] The groups `Vascular plants`, `Trees`, `Shrubs`, `Cloud Forest Trees`, `Endemic Trees`, `Perennial Shrubs`, `Arctic Vascular Plants`, `Spermatophytes` and `Angiosperms`, should have `Tracheophyta` as phylum.  
  - [ ] The group `Flowering Plant` should have 
  - [ ] The group `Ferns` should have `Tracheophyta` and `Polypodiopsida`, as phylum and class.
  - [ ] The group `Conifers`, should have class `Pinopsida`.      
  - [ ] The group `Palms`, should have `Tracheophyta`, `Liliopsida`, `Arecales`, and `Arecaceae`, as phylum, class, order and family.
  - [ ] The group `Monocotyledons` should have `Tracheophyta` and `Liliopsida`, as phylum and class.  
  - [ ] There's a typo in `Endemic and Range-Restricted Vascular Plantss`. This should have `Tracheophyta` as phylum.  

FIX

```{r}
#| label: check-taxon-fields-fix
#| echo: true
#| eval: false
#| message: false
#| warning: false

raw_metadata %>% 
  mutate(kingdom = ifelse(group == 'Flora', 'Plantae', kingdom)) %>% 
  mutate(kingdom = ifelse(group == 'Fauna', 'Animalia', kingdom)) %>%  
  mutate(phylum = ifelse(group == 'Flora' | group == 'Fauna', NA, phylum)) %>% 
  mutate(phylum = ifelse(group =='Onychophora', 'Onychophora', phylum)) %>% 
  mutate(group = ifelse(group =='Opilioness', 'Opiliones', group)) %>% 
  mutate(group = ifelse(group =='False scorpions', 'Pseudoscorpiones', group)) %>%
  mutate(order = ifelse(group =='False scorpions', 'Pseudoscorpiones', order)) %>%
  mutate(class = ifelse(group =='Entomostraca', NA, class)) %>%
  mutate(class = ifelse(group =='Horseshoe Crabs', 'Merostomata', class)) %>%
  mutate(class = ifelse(group =='Freshwater Crabs', 'Malacostraca', class)) %>%
  mutate(class = ifelse(group =='Sharks', 'Chondrichthyes', class)) %>%
  mutate(order = ifelse(iso_2 == 'KE' & group == 'Carnivores', 'Carnivora', order),
         order = ifelse(iso_2 == 'KE' & group == 'Perissodactyla', 'Perissodactyla', order),
         order = ifelse(iso_2 == 'KE' & group == 'Marine Cetartiodactyla', 'Artiodactyla', order),
         order = ifelse(iso_2 == 'KE' & group == 'Terrestrial Cetartiodactyla', 'Artiodactyla', order)) %>%
  mutate(group = ifelse(group == 'Marine Cetartiodactyla', 'Cetaceans', group),
         group = ifelse(group == 'Terrestrial Cetartiodactyla', 'Ungulates', group)) %>% 
  mutate(order = ifelse(group =='Endemic Lizards', 'Squamata', order)) %>%
  mutate(class = ifelse(grepl('lamprey', ignore.case=T, group), 'Petromyzonti', class),
         order = ifelse(grepl('lamprey', ignore.case=T, group), 'Petromyzontiformes', order)) %>%
  mutate(group = ifelse(group == 'Musslels', 'Mussels', group)) %>% 
  mutate(phylum = ifelse(grepl('orchid', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('orchid', ignore.case=T, group), 'Liliopsida', class),
         order = ifelse(grepl('orchid', ignore.case=T, group), 'Asparagales', order)) %>%
  mutate(phylum = ifelse(grepl('bromeli', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('bromeli', ignore.case=T, group), 'Liliopsida', class)) %>%
  mutate(phylum = ifelse(grepl('cinnamon', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('cinnamon', ignore.case=T, group), 'Magnoliopsida', class),
         order = ifelse(grepl('cinnamon', ignore.case=T, group), 'Canellales', order)) %>%
  mutate(phylum = ifelse(grepl('tree|shrub|vascular|angio|spermato|flower', ignore.case=T, group), 
                         'Tracheophyta', phylum)) %>% 
  mutate(phylum = ifelse(grepl('ferns', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('ferns', ignore.case=T, group), 'Polypodiopsida', class)) %>% 
  mutate(phylum = ifelse(grepl('conif', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('conif', ignore.case=T, group), 'Pinopsida', class)) %>% 
  mutate(phylum = ifelse(grepl('palm', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('palm', ignore.case=T, group), 'Liliopsida', class),
         order = ifelse(grepl('palm', ignore.case=T, group), 'Arecales', order)) %>%
  mutate(class = ifelse(grepl('monocot', ignore.case=T, group), 'Tracheophyta', class)) 
  
  
  
 
  
  


```

### Check `Event` fields

The fields is: `year`

  - Check it has numeric values.
  
```{r}
#| label: check-event-fields
#| echo: true
#| eval: false
#| message: false
#| warning: false


raw_metadata %>% 
  janitor::clean_names() %>% 
  janitor::remove_empty(c('rows', 'cols')) %>% 
  mutate(year = ifelse(year == 'NA', NA, year)) %>%
  mutate(year = ifelse(year == '2024?', NA, year)) %>%
  mutate(year = ifelse(name_orig == 'The Red List of Mammals of South Africa, Swaziland and Lesotho 2024', 
                       2024, year)) %>%
  mutate(year = as.numeric(year, na.rm=T)) %>% 
  filter(is.na(year)) %>% 
  select(year, name_orig) %>% 
  distinct(name_orig, .keep_all = T) 

```


#### DOUBTS

  - [ ] There are 16 records with year = `2024?`. Is it correct to change them for `2024`? 
  They belong to the sources:

```
https://www.odonat-grandest.fr/listes-rouges-grand-est-etat-avancement/
A Red List of Benin’s sharks                                           
The Red List of Mammals of South Africa, Swaziland and Lesotho 2024  
```

  - [ ] There are 9 records without year. What should we do with them? 
  They belong to the following source:
  
```
1 Crveni popis hrvatskih koralja                                                        
2 Crveni popis lišajeva Hrvatske                                                        
3 Červené seznamy                                                                       
4 Coleoptera (Beetle) – Invertebrate Ireland Online                                     
5 Tricoptera (Caddisfly) – InvertebrateIreland Online                                   
6 Les mammifères de la Côte d’Ivoire                                                    
7 Rongeurs et insectivores de Côte d’Ivoire, leur habitat et leur statut de conservation
8 Красная книга Азербайджанской Республики                                              
9 Красная книга Узбекистана  
```

### Check if URLs are working

```{r}
#| label: check-urls
#| echo: true
#| eval: true
#| message: false
#| warning: false

# URL incorrect
raw_metadata %>% 
  filter(!grepl('http', url_clean)) %>% 
  distinct(url_clean)

# URL error
# raw_metadata %>% filter(grepl('http', url_clean)) %>% 
#   distinct(url_clean) %>% 
#   mutate(check_URL = ifelse(map(URLencode(url_clean), http_error), 'not found', 'OK')) %>% 
#   filter(check_URL == 'not found') 

raw_metadata %>% 
  filter(!grepl('http', url_clean)) %>% 
  distinct(url_clean)

raw_metadata %>% 
  mutate(url_clean = ifelse(grepl('Nicolau', url_clean), NA, url_clean)) %>% 
  filter(!grepl('http', url_clean)) %>% 
  distinct(url_clean)
```

#### DOUBTS

  - [ ] There's one URL that resolves but has 'http' missing and one URL that is incorrect. Should this be `NA`?
  
```
ima.sc.gov.br/index.php/biodiversidade/biodiversidade/fauna
Nicolau, J. i Dalmau, J., 2008. Llista Vermella\r\ndels Vertebrats d’Andorra. BIOCOM (Biologia i\r\nComunicació) SL i Departament de Patrimoni\r\nNatural del Govern d’Andorra. Informe inèdit
```
  
## Run code and keep relevant fields

```{r}
#| label: run-code-and-keep-relevant-fields
#| echo: true
#| eval: true
#| message: false
#| warning: false

metadata <- 
  # check taxon
  left_join(raw_metadata %>% mutate(group = str_trim(group)) %>%
              select(-c(kingdom,phylum,subphylum,class,order)),
            bind_rows(merged_list %>% filter(!is.na(scientificName)),
                      merged_list %>% filter(is.na(scientificName)) %>%
                        select(group) %>%
                        left_join(. , raw_metadata_taxon_list))) %>% 
  mutate(group = str_trim(group)) %>% 
  mutate(group = str_to_title(group)) %>% 
  mutate(group = str_replace_all(group, 'And ', 'and ')) %>% 
  mutate(group = str_replace_all(group, 'Of', 'of')) %>% 
  mutate(group = str_replace_all(group, 'The', 'the')) %>%
  # more taxonomic corrections
  mutate(kingdom = case_when(grepl('flora', group, ignore.case=T) ~ 'Plantae',
                             grepl('fauna', group, ignore.case=T) ~ 'Animalia',
                             grepl('tunicata', group, ignore.case=T) ~ NA,
                             .default = kingdom)) %>% 
  mutate(phylum = ifelse(phylum == 'chordata', 'Chordata', phylum)) %>% 
  mutate(kingdom = ifelse(group == 'Flora', 'Plantae', kingdom)) %>% 
  mutate(kingdom = ifelse(group == 'Fauna', 'Animalia', kingdom)) %>%  
  mutate(phylum = ifelse(group == 'Flora' | group == 'Fauna', NA, phylum)) %>% 
  mutate(phylum = ifelse(group =='Onychophora', 'Onychophora', phylum)) %>% 
  mutate(group = ifelse(group =='Opilioness', 'Opiliones', group)) %>% 
  mutate(group = ifelse(group =='False scorpions', 'Pseudoscorpiones', group)) %>%
  mutate(order = ifelse(group =='False scorpions', 'Pseudoscorpiones', order)) %>%
  mutate(class = ifelse(group =='Entomostraca', NA, class)) %>%
  mutate(class = ifelse(group =='Horseshoe Crabs', 'Merostomata', class)) %>%
  mutate(class = ifelse(group =='Freshwater Crabs', 'Malacostraca', class)) %>%
  mutate(class = ifelse(group =='Sharks', 'Chondrichthyes', class)) %>%
  mutate(order = ifelse(iso_2 == 'KE' & group == 'Carnivores', 'Carnivora', order),
         order = ifelse(iso_2 == 'KE' & group == 'Perissodactyla', 'Perissodactyla', order),
         order = ifelse(iso_2 == 'KE' & group == 'Marine Cetartiodactyla', 'Artiodactyla', order),
         order = ifelse(iso_2 == 'KE' & group == 'Terrestrial Cetartiodactyla', 'Artiodactyla', order)) %>%
  mutate(group = ifelse(group == 'Marine Cetartiodactyla', 'Cetaceans', group),
         group = ifelse(group == 'Terrestrial Cetartiodactyla', 'Ungulates', group)) %>% 
  mutate(order = ifelse(group =='Endemic Lizards', 'Squamata', order)) %>%
  mutate(class = ifelse(grepl('lamprey', ignore.case=T, group), 'Petromyzonti', class),
         order = ifelse(grepl('lamprey', ignore.case=T, group), 'Petromyzontiformes', order)) %>%
  mutate(group = ifelse(group == 'Musslels', 'Mussels', group)) %>% 
  mutate(phylum = ifelse(grepl('orchid', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('orchid', ignore.case=T, group), 'Liliopsida', class),
         order = ifelse(grepl('orchid', ignore.case=T, group), 'Asparagales', order)) %>%
  mutate(phylum = ifelse(grepl('bromeli', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('bromeli', ignore.case=T, group), 'Liliopsida', class)) %>%
  mutate(phylum = ifelse(grepl('cinnamon', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('cinnamon', ignore.case=T, group), 'Magnoliopsida', class),
         order = ifelse(grepl('cinnamon', ignore.case=T, group), 'Canellales', order)) %>%
  mutate(phylum = ifelse(grepl('tree|shrub|vascular|angio|spermato|flower', ignore.case=T, group), 
                         'Tracheophyta', phylum)) %>% 
  mutate(phylum = ifelse(grepl('ferns', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('ferns', ignore.case=T, group), 'Polypodiopsida', class)) %>% 
  mutate(phylum = ifelse(grepl('conif', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('conif', ignore.case=T, group), 'Pinopsida', class)) %>% 
  mutate(phylum = ifelse(grepl('palm', ignore.case=T, group), 'Tracheophyta', phylum),
         class = ifelse(grepl('palm', ignore.case=T, group), 'Liliopsida', class),
         order = ifelse(grepl('palm', ignore.case=T, group), 'Arecales', order)) %>%
  mutate(class = ifelse(grepl('monocot', ignore.case=T, group), 'Tracheophyta', class)) %>% 
  # check columns
  janitor::clean_names() %>% 
  janitor::remove_empty(c('rows', 'cols')) %>% 
  # check source
  mutate(name_orig = str_squish(name_orig)) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\\\\")) %>% 
  mutate(name_orig = str_remove_all(name_orig, "\"")) %>% 
  # missing name of source
  mutate(name_orig = case_when(grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Birds' ~ 
                                 'Liste rouge des Oiseaux du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Mammals' ~ 
                                 'Liste rouge des Mammifères du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Orthoptera' ~ 
                                 'Liste rouge des Orthoptères du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Fishes' ~ 
                                 'Liste rouge des Papillons de jour (Rhopalocères et Zygènes) du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Butterflies' ~ 
                                 'Liste rouge des Papillons de nuit du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Night butterflies' ~ 
                                 'Liste rouge des Hétérocères du Grand Est',
                               grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Ladybugs' ~ 
                                 'Liste rouge des Coccinelles du Grand Est',
                                grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Branchiopoda' ~ 
                                 'Liste rouge des Branchiopodes du Grand Est',
                                grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Heteroptera' ~ 
                                 'Liste rouge des Punaises du Grand Est',
                                grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Mayflies' ~ 
                                 'Liste rouge des Ephémères – Trichoptères – Plécoptères du Grand Est',
                                grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Caddisflies' ~ 
                                 'Liste rouge des Ephémères – Trichoptères – Plécoptères du Grand Est',
                                grepl('htt', name_orig) & 
                               state_province == 'Grand Est' & group == 'Stoneflies' ~ 
                                 'Liste rouge des Ephémères – Trichoptères – Plécoptères du Grand Est',
                               .default = name_orig)) %>% 
  # check format
  mutate(format = ifelse(format == 'NA', NA, str_squish(format))) %>% 
  mutate(format = ifelse(format == '?', NA, format)) %>% 
  # check language
  mutate(language = ifelse(language == 'NA', NA, str_squish(language))) %>% 
  mutate(language = str_replace(language, "/|\\|", ' | ')) %>% 
  # check location
  mutate(continent = str_squish(str_replace_all(str_squish(continent), '\\|', ' | '))) %>%
  mutate(continent = str_replace_all(continent, '_', ' ')) %>%
  mutate(continent = str_to_title(continent)) %>% 
  mutate(country = ifelse(country == 'NA', NA, str_squish(country))) %>% 
  mutate(country = str_replace_all(country, '_', ' ')) %>%
  mutate(country = ifelse(country == 'USSR', NA, str_to_title(country))) %>%
  mutate(country = str_replace_all(country, 'And ', 'and ')) %>% 
  mutate(country = str_replace_all(country, 'Of', 'of')) %>% 
  mutate(country = str_replace_all(country, 'The', 'the')) %>%
  mutate(state_province = ifelse(state_province == 'NA', NA, str_squish(state_province))) %>%
  mutate(state_province = str_to_title(state_province)) %>% 
  mutate(iso_2 = ifelse(iso_2 == 'NA' & country != 'Namibia', NA, str_squish(iso_2))) %>% 
  mutate(iso_3 = ifelse(iso_3 == 'NA', NA, str_squish(iso_3))) %>%  
  mutate(iso_2 = str_squish(str_replace_all(str_squish(iso_2), '\\|', ' | '))) %>%
  mutate(iso_3 = str_squish(str_replace_all(str_squish(iso_3), '\\|', ' | '))) %>%
  mutate(gadm_level_1 = ifelse(gadm_level_1 == 'NA', NA, str_squish(gadm_level_1))) %>%
  mutate(gadm_level_1 = str_to_title(gadm_level_1)) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'And ', 'and ')) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'Of', 'of')) %>% 
  mutate(gadm_level_1 = str_replace_all(gadm_level_1, 'The', 'the')) %>% 
  mutate(gadm_level_2 = ifelse(gadm_level_2 == 'NA', NA, str_squish(gadm_level_2))) %>%
  mutate(gadm_level_2 = str_to_title(gadm_level_2)) %>% 
  mutate(region_custom = ifelse(region_custom == 'NA', NA, str_squish(region_custom))) %>% 
  mutate(region_detail = ifelse(region_detail == 'NA', NA, str_squish(region_detail))) %>% 
  mutate(region_detail = str_squish(str_replace_all(str_squish(region_detail), '\\|', ' | '))) %>%
  # check event
  mutate(year = ifelse(year == 'NA', NA, year)) %>%
  mutate(year = ifelse(year == '2024?', NA, year)) %>%
  mutate(year = ifelse(name_orig == 'The Red List of Mammals of South Africa, Swaziland and Lesotho 2024', 
                       2024, year)) %>%
  mutate(year = as.numeric(year, na.rm=T)) %>% 
  # check urls
  mutate(url_clean = ifelse(url_clean == 'NA', NA, url_clean)) %>% 
  mutate(url_clean = ifelse(grepl('Nicolau', url_clean), NA, url_clean)) %>% 
  # select columns
  select(id, continent, 
         gadm_level_0 = country, gadm_level_1, gadm_level_2,
         region_custom, region_detail, iso_2, iso_3,
         taxa=group, kingdom, phylum, class, order, family, 
          source_name= name_orig, source_link = url_clean,
         language, year) 

metadata %>% slice_sample(n=50) %>% arrange(id) %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))
```


## Data audit

### Summary

```{r}
#| label: summary
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true

metadata %>% 
  summarise(`Number of records` = n(),
            `Number of sources` = n_distinct(source_name),
            `Number of taxa` = n_distinct(taxa),
            `Animalia records` = sum(kingdom == 'Animalia'),
            `Plantae records` = sum(kingdom == 'Plantae'),
            `Fungi records` = sum(kingdom == 'Fungi'),
            Countries = n_distinct(gadm_level_0),
            `Sub-national territories` = n_distinct(gadm_level_1)) %>% 
  t() %>% `colnames<-`(c("N")) %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))
```


### Geographic coverage

```{r}
#| label: geo-coverage
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true

metadata %>%
  separate_rows(continent, sep = '\\|') %>%
  mutate(continent = str_squish(continent)) %>%
  group_by(continent) %>% 
  summarise(n_sources = ifelse(n_distinct(source_name, na.rm = TRUE)==0, 
                               0, n_distinct(source_name, na.rm = TRUE))) %>% 
  arrange(desc(n_sources)) %>% 
  rename(`Number of sources`=n_sources) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position')) 

metadata %>%
  separate_rows(continent, sep = '\\|') %>%
  mutate(continent = str_squish(continent)) %>%
  group_by(continent) %>% 
  summarise(n_records = n()) %>% 
  arrange(desc(n_records)) %>% 
  rename(`Number of records`=n_records) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position')) 

# Europe
metadata %>%
  separate_rows(continent, sep = '\\|') %>%
  mutate(continent = str_squish(continent)) %>%
  filter(continent == 'Europe') %>% 
  group_by(kingdom) %>% 
  summarise(n_records = n()) %>% 
  arrange(desc(n_records)) %>% 
  rename(`Number of records`=n_records) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position')) 

metadata %>%
  separate_rows(continent, sep = '\\|') %>%
  mutate(continent = str_squish(continent)) %>%
  filter(continent == 'Asia') %>% 
  group_by(kingdom) %>% 
  summarise(n_records = n()) %>% 
  arrange(desc(n_records)) %>% 
  rename(`Number of records`=n_records) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))


world <- geodata::world(resolution = 3, level = 0, path = 'data/')

world_records <- left_join(st_as_sf(world), metadata %>%
  separate_rows(iso_3, sep = '\\|') %>% 
  select(iso_2, GID_0=iso_3, source_name)) %>% 
  group_by(GID_0, NAME_0) %>% 
  summarise(n_sources = ifelse(n_distinct(source_name, na.rm = TRUE)==0, 
                               0, n_distinct(source_name, na.rm = TRUE)),
            iso_2_string = ifelse(n_sources>0,
                                  paste(iso_2, collapse = ';'), NA)) %>% 
  ungroup() %>% st_cast() %>% st_set_crs(4326)

plot_figure_1 <- tm_shape(world_records %>% 
           select(-iso_2_string) %>% 
           mutate(n_sources=ifelse(n_sources==0,
                                   NA, n_sources))) +
  tm_polygons(fill = 'n_sources',fill_alpha = 0.9,
              col='grey40', col_alpha = 0.2,
              fill.scale = tm_scale_intervals(n = 6, 
                                              #style = 'jenks', 
                                              breaks = c(1,5,10,20,100,979),
                                              values = 'brewer.reds',
                                              value.na = 'grey80',
                                              label.na = '0'),
              fill.legend = tm_legend(item.space = 0, item.na.space = 0,
                                      title = 'Number of sources', 
                                      reverse=T, 
                                      # frame=F, 
                                      frame.lwd = 0.1,
                                      bg.color = 'white')) +
  tm_layout(legend.outside = T, 
            legend.position = c('left','bottom'),  frame=F) +
  tm_crs(property='global')

tmap_mode('plot')
plot_figure_1
tmap_mode('view')
plot_figure_1
```

### Taxonomic coverage

```{r}
#| label: taxonomic-coverage
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true

# by kingdom
metadata %>%
  group_by(kingdom) %>% 
  summarise(n_records = n()) %>% 
  arrange(desc(n_records)) %>% 
  rename(`Number of records` = n_records) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))

# animalia
metadata %>% 
    filter(!is.na(class)) %>% 
    filter(kingdom %in% c('Animalia')) %>% 
    group_by(kingdom, class) %>% 
    summarise(n_sources_taxa = n()) %>% 
    arrange(desc(n_sources_taxa)) %>% 
    slice_head(n=5) %>% 
  rename(`Number of sources` = n_sources_taxa) %>% 
  adorn_totals('row') %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))

# plantae
metadata %>% 
    filter(!is.na(order)) %>% 
    filter(kingdom %in% c('Plantae')) %>% 
    group_by(kingdom, order) %>% 
    summarise(n_sources_taxa = n_distinct(source_name)) %>% 
    arrange(desc(n_sources_taxa)) %>% 
    slice_head(n=5) %>% 
  rename(`Number of sources` = n_sources_taxa) %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))

# fish sources
metadata %>% 
  filter(grepl('fish', taxa, ignore.case=T)) %>% 
  filter(!grepl('crayfish', taxa, ignore.case=T)) %>% 
  distinct(source_name) %>% count() %>% 
  rename(`Number of fish sources` = n) %>% 
  kableExtra::kbl(booktabs = T)  %>% 
  kableExtra::kable_styling(latex_options = c('striped', 'hold_position'))

metadata %>% 
  filter(!is.na(phylum)) %>% 
  filter(kingdom %in% c('Animalia')) %>% 
  group_by(kingdom, phylum) %>% 
  summarise(n_sources_taxa = n_distinct(source_name)) %>% 
  ggplot(aes(x = reorder(phylum, -n_sources_taxa), y = n_sources_taxa)) +
  geom_bar(stat = "identity", fill = "#4CAF50") +
  facet_wrap(~kingdom, scales = 'free') +
  labs(x = "", y = "Number of sources") +
  ggpubr::theme_pubclean() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

metadata %>% 
  filter(!is.na(phylum)) %>% 
  filter(kingdom %in% c('Plantae', 'Fungi')) %>% 
  group_by(kingdom, phylum) %>% 
  summarise(n_sources_taxa = n_distinct(source_name)) %>% 
  ggplot(aes(x = reorder(phylum, -n_sources_taxa), y = n_sources_taxa)) +
  geom_bar(stat = "identity", fill = "#4CAF50") +
  facet_wrap(~kingdom, scales = 'free') +
  labs(x = "", y = "Number of sources") +
  ggpubr::theme_pubclean() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

metadata %>% 
  filter(!is.na(order)) %>% 
  filter(kingdom %in% c('Animalia')) %>% 
  group_by(kingdom, order) %>% 
  summarise(n_sources_taxa = n_distinct(source_name)) %>% 
  ggplot(aes(x = reorder(order, -n_sources_taxa), y = n_sources_taxa)) +
  geom_bar(stat = "identity", fill = "#4CAF50") +
  facet_wrap(~kingdom, scales = 'free') +
  labs(x = "", y = "Number of sources") +
  ggpubr::theme_pubclean() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

metadata %>% 
  filter(!is.na(order)) %>% 
  filter(kingdom %in% c('Plantae', 'Fungi')) %>% 
  group_by(kingdom, order) %>% 
  summarise(n_sources_taxa = n_distinct(source_name)) %>% 
  ggplot(aes(x = reorder(order, -n_sources_taxa), y = n_sources_taxa)) +
  geom_bar(stat = "identity", fill = "#4CAF50") +
  facet_wrap(~kingdom, scales = 'free') +
  labs(x = "", y = "Number of sources") +
  ggpubr::theme_pubclean() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

### Temporal coverage

```{r}
#| label: temporal-coverage
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true

metadata %>% 
  group_by(year) %>% 
  summarise(publications_year = n_distinct(source_name)) %>% 
  ggplot(aes(x = year, y = publications_year)) +
  geom_bar(stat = "identity", fill = "#4CAF50") +
  ylim(c(0,150)) + xlim(c(1975,2025)) +
  scale_x_continuous(n.breaks = 15) +
  labs(x = "", y = "Number of sources") +
  ggpubr::theme_pubclean()
```

## Save file

```{r}
#| label: save-file
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true

write_csv(metadata, 'data/metadata.csv', na = '')
```
